---
title: "Full blood count analysis"
author: "Harrison Wood"
date: "9/19/2019"
output: html_document
---

```{r Install packages, eval=FALSE, include=FALSE}
#Install the excel file reader package
install.packages("readxl")

#Install the RColorBrewer package
install.packages("RColorBrewer")

#Install the rglwidget package for 3D plots
install.packages("rglwidget")

#Install the kableExtra package for styled tables
install.packages("kableExtra")

#Install the condformat package for table based colour keys
install.packages("condformat")

#Install the formattable and datatable packages for presentation of results
install.packages("formattable")
install.packages("DT")

#Install the nonlinear mixed effects models package for statistical modelling
install.packages("nlme")
```

```{r Setup, include=FALSE}
#Load knitr library
library(knitr)

#Set initial knitr options
knitr::opts_knit$set(
  fig.width = 12,
  fig.height = 8,
  warning = FALSE,
  message = FALSE,
  root.dir = normalizePath(".")
)

#Load libraries for 3D plots
library(rgl)
library(rglwidget)
#Load library for reading Excel files
library(readxl)
#Load library for making colour schemes
library(RColorBrewer)
#Load library for biplots
library(ggbiplot)
#Load library for styled kable tables
library(kableExtra)
#Load library for table based colour keys
library(condformat)
#Load libraries for tabular presentation of results
library(formattable)
library(DT)
#Load library for nonlinear statistical modelling
library(nlme)

#Set up rgl hooks for 3D plots in Knitr
setupKnitr()
```

## Loading data

Data are stored in the ./data/ folder, and consist of full blood count measurements from an Advia 2120i Haematology system. Measurements were made on blood draws obtained via cardiac puncture of isofluorane-anaesthetised mice and stored in EDTA-coated tubes. 

First we load the data in .xlsx format

```{r Load data}
#Load the summary blood data
rawData <- read_excel("data/Summary-blood-data.xlsx")
data <- as.data.frame(rawData)
```

```{r Show data, eval=TRUE, echo=FALSE}
#We can then inspect the basic structure of the dataset
as.datatable(formattable(data),options = list(scrollX = TRUE))
```

## Clean up data

We then clean up the data for further analysis. This includes removing labels and descriptors, failed samples, and, optionally, categorical descriptors.

```{r Clean up data}
#Remove failed samples
removeFailures <- TRUE
if (removeFailures) {
  completedSamples <- data$`Processed?` == "Y"
  data <- data[completedSamples,]
}

#Remove abnormal samples
removeAbnormals <- TRUE
if (removeAbnormals) {
  normalSamples <- data$atyps_cond == 0
  data <- data[normalSamples,]
}

#Clean up the data by excluding fields irrelevant to the analysis
labelFields <-
  c(
    "Age",
    "Processed?",
    "Notes",
    "Idee",
    "Instr#",
    "SampleMode",
    "SampleSel",
    "Sequence",
    "Date/Time"
  )
data <- data[, !(names(data) %in% labelFields)]

#Drop the fields containing categorical descriptions
categoricalFields <-
  c(
    "micro_cond",
    "macro_cond",
    "hypo_cond",
    "hyper_cond",
    "aniso_cond",
    "var_cond",
    "left_shift_cond",
    "blast_cond",
    "atyps_cond",
    "nrbc_cond",
    "ig_cond",
    "large_plt_cond",
    "plt_clumps_cond",
    "mpo_deficiency",
    "rbc_fragments_condition",
    "rbc_ghosts_condition"
  )
removeCategoricals <- TRUE
if (removeCategoricals) {
  data <- data[, !(names(data) %in% categoricalFields)]
}

#Clean up variable names to permit use in formulae later in analysis
names(data)<-make.names(names(data))

#Ensure each animal has a unique identifier
data[,"ID.."]<-make.unique(as.character(data[,"ID.."]))
```

## Experimental design

We then factorise the independent variables and set up descriptors for the experimental design to prepare for plotting and later analysis.

```{r Experimental design}
#Group genotypes based on Actn3 presence or absence
data$Actn3 <- factor((data$Genotype == "WT" | data$Genotype == "RR"))
levels(data$Actn3) <- c("KO", "WT")

#Factor independent variables describing the experimental design
independentVariables <-
  c("ID..",
    "Model",
    "Genotype",
    "Sex",
    "Intervention",
    "Date.collected",
    "Actn3")
data[independentVariables] <-
  lapply(data[independentVariables], factor)

#Generate a list of the dependent properties being measured
dependentVariables <- colnames(data)[!colnames(data)%in%independentVariables]

#Define colours for each grouping of independent variables
#Generate a data frame containing the colour key for the independent variable to be assessed
generateColourKey <- function(plotGrouping = plotGrouping) {
  plotColours <-
    suppressWarnings(I(brewer.pal(nlevels(data[, plotGrouping]), name =
                                    "Set1"))[1:nlevels(data[, plotGrouping])])
  names(plotColours) <- levels(data[, plotGrouping])
  return(plotColours)
}

#Generate a colour legend for the independent variable
showColourLegend <- function(plotColours = plotColours) {
  suppressWarnings(
    condformat(data.frame(Category = names(plotColours))) +
      rule_fill_discrete(1, expression = names(plotColours), colours = plotColours)
  )
}
```

## Principal component analysis

We then calculate the principal components of the dataset after centering and scaling, and plot the PCA coloured by each experimental factor to investigate the distrubtion of samples.

```{r PCA, echo=TRUE}
#Exclude outlier samples biasing the PCA
outlierSamples <- c("313")
excludeOutliers <- TRUE
if (excludeOutliers) {
  #Filter out the outliers
  PCAdata <- data[!data$ID.. %in% outlierSamples, ]
}

#Extract the labels
PCAlabels<-PCAdata[, (names(PCAdata) %in% independentVariables)]
#Then trim them from the dataset
PCAdata<-PCAdata[, !(names(PCAdata) %in% independentVariables)]

#Compute the PCA
PCA<-prcomp(PCAdata,center=TRUE,scale=TRUE)
```

```{r Define PCA plotting functions, include=FALSE}
#Plot the PCA along the given components
plot3DPCA <- function(PCA = PCA,
                    components = c(1, 2, 3),
                    plotGrouping = plotGrouping,
                    plotColours = plotColours) {
  plot3d(
    PCA$x[, components[1]],
    PCA$x[, components[2]],
    PCA$x[, components[3]],
    xlab = paste(
      "PC ",
      components[1],
      " (",
      round(summary(PCA)$importance[2, components[1]] * 100, 1),
      "% var.)",
      sep =
        ""
    ),
    ylab = paste(
      "PC ",
      components[2],
      " (",
      round(summary(PCA)$importance[2, components[2]] * 100, 1),
      "% var.)",
      sep =
        ""
    ),
    zlab = paste(
      "PC ",
      components[3],
      " (",
      round(summary(PCA)$importance[2, components[3]] * 100, 1),
      "% var.)",
      sep =
        ""
    ),
    type = "p",
    size = 6,
    col = plotColours[match(PCAlabels[, plotGrouping], names(plotColours))]
  )
  title3d(paste("PCA coloured by ", plotGrouping, sep = ""))
  bg3d("grey90")
}

#Plot two parallel PCAs for a given variable
graphVariable <- function(variable = plotGrouping, PCA = PCA, plotColours=plotColours) {
  mfrow3d(1, 2)
  plot3DPCA(
    PCA = PCA,
    components = c(1, 2, 3),
    plotGrouping = variable,
    plotColours = plotColours
  )
  plot3DPCA(
    PCA = PCA,
    components = c(1, 4, 5),
    plotGrouping = variable,
    plotColours = plotColours
  )
}
```

### Plot each independent variable
Click and drag on any plot to rotate the view.

### Actn3 
```{r Plot Actn3, webgl=TRUE, message=FALSE, echo=FALSE, results="asis", fig.width=9, fig.height=6}
plotGrouping <- "Actn3"
plotColours<-generateColourKey(plotGrouping)
showColourLegend(plotColours)
graphVariable(variable=plotGrouping,PCA=PCA,plotColours=plotColours)
```

### Sex 
```{r Plot Sex, webgl=TRUE, message=FALSE, echo=FALSE, results="asis", fig.width=9, fig.height=6}
plotGrouping <- "Sex"
plotColours<-generateColourKey(plotGrouping)
showColourLegend(plotColours)
graphVariable(variable=plotGrouping,PCA=PCA,plotColours=plotColours)
```

### Genotype 
```{r Plot Genotype, webgl=TRUE, message=FALSE, echo=FALSE, results="asis", fig.width=9, fig.height=6}
plotGrouping <- "Genotype"
plotColours<-generateColourKey(plotGrouping)
showColourLegend(plotColours)
graphVariable(variable=plotGrouping,PCA=PCA,plotColours=plotColours)
```

### Model 
```{r Plot Model, webgl=TRUE, message=FALSE, echo=FALSE, results="asis", fig.width=9, fig.height=6}
plotGrouping <- "Model"
plotColours<-generateColourKey(plotGrouping)
showColourLegend(plotColours)
graphVariable(variable=plotGrouping,PCA=PCA,plotColours=plotColours)
```

### Intervention 
```{r Plot Intervention, webgl=TRUE, message=FALSE, echo=FALSE, results="asis", fig.width=9, fig.height=6}
plotGrouping <- "Intervention"
plotColours<-generateColourKey(plotGrouping)
showColourLegend(plotColours)
graphVariable(variable=plotGrouping,PCA=PCA,plotColours=plotColours)
```

### Sex effect biplot

We observe an apparent delineation by sex along PCs 1 and 5, and can plot this with the loadings for further investigation.

```{r Biplot, eval=TRUE}
plotGrouping <- "Sex"
plotColours <- generateColourKey(plotGrouping)
colScale <-
  scale_colour_manual(name = "Category", values = plotColours)
ggbiplot(
  PCA,
  choices = c(1, 5),
  labels = PCAlabels$ID..,
  groups = PCAlabels[, plotGrouping],
  ellipse = TRUE,
) + colScale
```

## Statistical modelling

Testing the effects of the experimental factors is done by estimating parameters for a linear mixed effects model, as below. 

```{r General linear model, eval=TRUE}
#Define which parameters should be tested
testParameters <- dependentVariables

#Define the model to test
specifiedModel <-
  "~Intervention+Actn3+Model+Sex+(Model:Actn3)+(Sex:Actn3)"
numberOfModelParameters <- 7

#Define our threshold for significance (highlights p values later)
sigThreshold <- 0.05

#Define function to estimate a model for a given parameter
calculateModel <-
  function(modelDataset = data,
           modelToTest = specifiedModel,
           parameter = dependentVariables[1]) {
    lmm <-
      lme(
        fixed = as.formula(paste(parameter,
                                 modelToTest,
                                 sep = "")),
        random =  ~ 1 | Date.collected,
        method = "REML",
        data = modelDataset
      )
    return(lmm)
  }

#Set numerical display options, then show results
options(digits = 5, scipen = 15)

#Set up data frame to store model results
modelResults <-
  data.frame(
    matrix(
      vector(),
      numberOfModelParameters,
      length(testParameters),
      dimnames = list(c(), testParameters)
    ),
    stringsAsFactors = F
  )

#Evaluate the model for each parameter to be tested and store the results
for (i in seq_along(testParameters)) {
  parameterResult <-
    data.frame(anova.lme(calculateModel(parameter = testParameters[i])), type = "marginal")
  modelResults[, testParameters[i]] <-
    round(parameterResult[, "p.value"], 5)
}
rownames(modelResults) <- rownames(parameterResult)

#Drop the intercept row of the results table
modelResults <- modelResults[!rownames(modelResults) == "(Intercept)", ]
#Transpose for readability
modelResults <- as.data.frame(t(modelResults))
```

### Coefficient p values
```{r Display results table, echo=FALSE}
as.datatable(formattable(modelResults, list(
  area(col = 1:6) ~ formatter("span", style = x ~ ifelse(
    x < sigThreshold,
    style(color = "red", font.weight = "bold"), NA
  ))
)))
```